---
title: "Red Wine Exploration by Justin Rogers"
output: html_document
---

## This report explores which chemical properties influence the quality of red ws.

# Data Exploration
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8)
```

```{r}
wines <- read.csv("wineQualityReds.csv")
# Remove the columns containing the index
ws <- wines[, c(2:13)]
library("GGally")
library("ggplot2")
library("dplyr")
library(gridExtra)
library(grid)
```

```{r}
dim(ws)
```

```{r}
str(ws)
```
```{r}
summary(ws)
```
The report explores a dataset containing quality and 11 features for 1599 red wines observations. 


```{r}
sapply(ws, function(x) sum(is.na(x)))
```
Looking for the number of NA values for each column in the dataframe. It appears that none are missing.

```{r}
cor(ws[,c(1:11)], ws[,c(12)])
```
Correlation showing all variables against quality. It appears that four attributes have a weak to moderate correlation (either negative or positive) with quality: volatile.acidity, citric.acid, sulphates, and alcohol.

```{r}
ggpairs(ws)
```
A quick matrix chart showing some of the relationships between variables.


```{r}
 ggplot(aes(x = quality), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```
```{r}
summary(ws$quality)
```
The qualities conform to a fairly normal distribution. While the scores limits were 0-10, no wines fell below 3 or scored above 8 and most falling below a 6.

Now we will review some of the attributes highly correlated with quality.
```{r}
ggplot(aes(x = alcohol), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```

```{r}
ggplot(aes(x = volatile.acidity), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```
```{r}
ggplot(aes(x = sulphates), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```

```{r}
ggplot(aes(x = citric.acid), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```

```{r}
ggplot(aes(x = quality, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  scale_y_continuous(limits = c(8,15), breaks = seq(8, 15, 1))
```

# Final Plots and Summary

# Reflections


