---
title: "Red Wine Exploration by Justin Rogers"
output: html_document
---

## This report explores which chemical properties influence the quality of red wines.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8)
```

```{r}
wines <- read.csv("wineQualityReds.csv")
# Remove the columns containing the index
ws <- wines[, c(2:13)]
library("GGally")
library("ggplot2")
library("dplyr")
library(gridExtra)
library(grid)
library(e1071)
library(RColorBrewer)
```

# Univariate Plots Section

The report explores a dataset containing quality and 11 features for 1599 red wines observations.

```{r}
dim(ws)
```

```{r}
str(ws)
```
```{r}
summary(ws)
```
 
```{r}
sapply(ws, function(x) sum(is.na(x)))
```
Looking for the number of NA values for each column in the dataframe. It appears that none are missing.

```{r}
cor(ws[,c(1:11)], ws[,c(12)])
```
Correlation showing all variables against quality. It appears that four attributes have a weak to moderate correlation (either negative or positive) with quality: volatile.acidity, citric.acid, sulphates, and alcohol.

```{r}
cor(ws)
```
Correlations between all variables.
  

```{r}
 ggplot(aes(x = quality), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
```

```{r}
summary(ws$quality)
```
The qualities conform to a fairly normal distribution. While the scores limits were 0-10, no wines fell below 3 or scored above 8 and most falling below a 6.

### Explore the four variables with the highest correlations with quality

```{r}
p1 = ggplot(aes(x = volatile.acidity), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 = ggplot(aes( x = 1, y = volatile.acidity ), data = ws) + 
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = "#099DD9")
grid.arrange(p1, p2, ncol=2)
```
```{r}
summary(ws$volatile.acidity)
```
```{r}
skewness(ws$volatile.acidity)
```
Volatile acidity is positively skewed.

```{r}
#Removing outliers to check on correlation
outliers <- boxplot.stats(ws$volatile.acidity)$out
ws_out <- ws
ws_out$outlier <- ifelse(ws_out$volatile.acidity %in% outliers, NA, ws_out$volatile.acidity)
dim(ws)
ws_out <- ws_out[!is.na(ws_out$outlier),]
dim(ws_out)
length(outliers)
cor(ws_out$volatile.acidity, ws_out$quality)
```
Removal of outliers did not improved the correlation with quality.

```{r}
p1 <- ggplot(aes(x = log10(volatile.acidity + 1)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 <- ggplot(aes(x = sqrt(volatile.acidity)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p3 <- ggplot(aes(x = sqrt(sqrt(volatile.acidity))), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p4 <- ggplot(aes(x = volatile.acidity^1/3), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p5 <- ggplot(aes(x = 1/volatile.acidity), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
grid.arrange(p1, p2, p3, p4, p5,  ncol = 1)
```
Squareroot of squareroot appears to give the best normal distribution.

```{r}
cor(sqrt(sqrt(ws$volatile.acidity)), ws$quality)
```
But it doesn't create a much stronger correlation.

```{r}
p1 = ggplot(aes(x = citric.acid), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 = ggplot(aes( x = 1, y = citric.acid ), data = ws) + 
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = "#099DD9")
grid.arrange(p1, p2, ncol=2)
```
```{r}
summary(ws$citric.acid)
```
```{r}
skewness(ws$citric.acid)
```
Citric acid appears to be positively skewed, but the jump around .5 reduces the skewness measure.

```{r}
outliers <- boxplot.stats(ws$citric.acid)$out
ws_out <- ws
ws_out$outlier <- ifelse(ws_out$citric.acid %in% outliers, NA, ws_out$citric.acid)
ws_out <- ws_out[!is.na(ws_out$outlier),]
cor(ws_out$citric.acid, ws_out$quality)
```
The removal of outliers did not improve correlation which makes sense since the chart doesn't appear to show any outliers.

```{r}
p1 <- ggplot(aes(x = log10(citric.acid + 1)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 <- ggplot(aes(x = sqrt(citric.acid)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p3 <- ggplot(aes(x = sqrt(sqrt(citric.acid))), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p4 <- ggplot(aes(x = citric.acid^1/3), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p5 <- ggplot(aes(x = 1/citric.acid), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
grid.arrange(p1, p2, p3, p4, p5,  ncol = 1)
```
Through all the transforms, it appears that squareroot creates the most normal distribution, but still has a large number of wines with almost no citric acid.
```{r}
cor(sqrt(ws$citric.acid), ws$quality)
```
And the squareroot actually lowers the correlation.

```{r}
p1 = ggplot(aes(x = sulphates), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 = ggplot(aes( x = 1, y = sulphates), data = ws) + 
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = "#099DD9")
grid.arrange(p1, p2, ncol=2)
```
```{r}
summary(ws$sulphates)
```
```{r}
skewness(ws$sulphates)
```
Sulphates are highly positively skewed.

```{r}
outliers <- boxplot.stats(ws$sulphates)$out
ws_out <- ws
ws_out$outlier <- ifelse(ws_out$sulphates %in% outliers, NA, ws_out$sulphates)
ws_out <- ws_out[!is.na(ws_out$outlier),]
cor(ws_out$sulphates, ws_out$quality)
```
Removing sulphate significantly improves correlation with quality from .25 to .39

```{r}
p1 <- ggplot(aes(x = log10(sulphates + 1)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 <- ggplot(aes(x = sqrt(sulphates)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p3 <- ggplot(aes(x = sqrt(sqrt(sulphates))), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p4 <- ggplot(aes(x = sulphates^1/3), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p5 <- ggplot(aes(x = 1/sulphates), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
grid.arrange(p1, p2, p3, p4, p5,  ncol = 1)
```
Reciprical transform shows the best normal distribution.
```{r}
cor(1/(ws$sulphates), ws$quality)
```
This actually increased correlation and turned it negative. We'll explore both options in the bivariate section.


```{r}
p1 = ggplot(aes(x = alcohol), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 = ggplot(aes( x = 1, y = alcohol), data = ws) + 
               geom_jitter(alpha = 0.1) +
               geom_boxplot(alpha = 0.2, color = "#099DD9")
grid.arrange(p1, p2, ncol=2)
```
```{r}
summary(ws$alcohol)
```
```{r}
skewness(ws$alcohol)
```
Alcohol has a positive skew.

```{r}
outliers <- boxplot.stats(ws$alcohol)$out
ws_out <- ws
ws_out$outlier <- ifelse(ws_out$alcohol %in% outliers, NA, ws_out$alcohol)
ws_out <- ws_out[!is.na(ws_out$outlier),]
cor(ws_out$alcohol, ws_out$quality)
```
Removing alcohol outliers did not improve correlation with quality.

```{r}
p1 <- ggplot(aes(x = log10(alcohol + 1)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p2 <- ggplot(aes(x = sqrt(alcohol)), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p3 <- ggplot(aes(x = sqrt(sqrt(alcohol))), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p4 <- ggplot(aes(x = alcohol^1/3), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
p5 <- ggplot(aes(x = 1/alcohol), data = ws) + 
  geom_histogram(color = I("black"), fill = I("#099DD9"))
grid.arrange(p1, p2, p3, p4, p5,  ncol = 1)
```
The best normalization is created by the squareroot transform.
```{r}
cor(sqrt(ws$alcohol), ws$quality)
```
No real change in corrlation, but we'll plot both in the bivariate section.

# Univariate Analysis

### What is the structure of your dataset?
There are 1,599 red wines in the dataset with 12 features (fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol and quality).

* Most red wines are of qualities 5 and 6.
* The median quality is 6.
* The mean alcohol is wines was 10.42.
* About 75% of red wines had sulphates less than .73.

### What is/are the main feature(s) of interest in your dataset?
The variables volatile acidity, citric acid, sulphates, and alcohol have the highest correlations with quality.

### Of the features you investigated, were there any unusual distributions? \
I checked for NA values and found none.

For the four features with the highest correlations, I removed outliers and reperformed correlations with quality. This only made a significant increase in correlation with sulphates.

I also performed log, squareroot, squareroot of squareroot, cube root, and reciprocal transforms for all four features with the highest correlation to determine if more normal distribution could be created and higher correlations with quality could be confirmed. Higher correlations were not achieved.

# Bivariate Plots Section

```{r}
ggpairs(ws)
```

```{r}
ggplot(aes(x = quality, y = volatile.acidity), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") +
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
The negative correlation is obvious in the trendline between volatile acidity and quality.

```{r}
ggplot(aes(x = quality, y = citric.acid), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") +
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
The weak positive correlation can be seen in the trendline between citric acid and quality.

```{r}
ggplot(aes(x = quality, y = sulphates), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") +
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
```{r}
ggplot(aes(x = quality, y = 1/(ws$sulphates)), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") +
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
The reciprocal sulphates show a much stronger negative trend with quality.

```{r}
ggplot(aes(x = quality, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
```{r}
ggplot(aes(x = quality, y = sqrt(alcohol)), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  scale_x_continuous(limits = c(2,9), breaks = seq(2, 9, 0.5)) +
  geom_smooth(method = "lm")
```
Both alchohol and squareroot transform appear to have the strongest trendlines we've seen with quality.

Now we'll see see how the four attributes with the hightest correlations with quality correlate with each other.
```{r}
ggplot(aes(x = volatile.acidity, y = citric.acid), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$volatile.acidity, ws$citric.acid)
```
Pretty strong negative correlation between volatile acidity and citric acid, but both of those attributes could be correlated merely because they are acids in the wine. 

```{r}
ggplot(aes(x = volatile.acidity, y = sulphates), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$volatile.acidity, ws$sulphates)
```
Weak correlation between volatile acidity and sulphates.

```{r}
ggplot(aes(x = volatile.acidity, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$volatile.acidity, ws$alcohol)
```
Weak correlation between volatile acidity and alcohol.

```{r}
ggplot(aes(x = citric.acid, y = sulphates), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$citric.acid, ws$sulphates)
```
Medium correlation between citric acid and sulphates.

```{r}
ggplot(aes(x = citric.acid, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$citric.acid, ws$alcohol)
```
Very little correlation between citric acid and alcohol.

```{r}
ggplot(aes(x = sulphates, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  geom_smooth(method = "lm")
```
```{r}
cor(ws$sulphates, ws$alcohol)
```
Very little correlation between sulphates and alchohol.

# Bivariate Analysis

Alcohol appears the be the highest correlation with quality followed by sulphates as far as positive correlations. Volatile acidity, citric acidity, and sulphates have medium correlations with each other.

The most interesting item is that positively correlated with each other, both citric acid and sulphates have very low correlations with alcohol. This points to a combination of either alcohol and citric acid or sulphates being excellent attributes to use in prediction.

# Multivariate Plots Section

```{r}
ggplot(aes(x = alcohol, y = residual.sugar, color = factor(quality)), data = red.wine) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
  guide=guide_legend(title='Quality'))
```

# Multivariate Analysis

# Final Plots and Summary
## Plot One
```{r}
ggplot(aes(x = quality, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  scale_x_continuous(limits = c(3,8.5), breaks = seq(3, 8.5, 0.5)) +
  scale_y_continuous(limits = c(9,13), breaks = seq(9, 13, 0.5)) +
  geom_smooth(method = "lm") +
  labs(title = "Quality by Alcohol Content") +
  xlab("Quality") +
  ylab("Alcohol (% by volume)") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Description One
Alcohol had the highest positive correlation with wine quality. This makes sense as one of the primary reason to have an alcoholic beverage in the first place is for alcohol. At around 7 level quality the vast majority of those wines contain an alcohol percentage greater than 10%.

## Plot Two
```{r}
ggplot(aes(x = quality, y = sulphates), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") +
  scale_x_continuous(limits = c(3,8), breaks = seq(3,8, 0.5)) +
  scale_y_continuous(limits = c(.3,1.25), breaks = seq(.3,1.25, 0.1)) +
  geom_smooth(method = "lm") +
  labs(title = "Quality by Sulphates") +
  xlab("Quality") +
  ylab("Sulphate Level (potassium sulphate - g / dm3)") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Description Two
Sulphates had the second highest positive correlation with quality. Sulphates are additives to wines which acts as antimicrobial and antioxidant agents. These preserve the wines so perhaps an increase in sulphates would produce less likelihood that the wine tasted would have gone bad.

## Plot Three
```{r}
ggplot(aes(x = sulphates, y = alcohol), data = ws) + 
  geom_point(alpha = 5/10, position = position_jitter(h = 0), color = "#099DD9") + 
  scale_x_continuous(limits = c(.3,1.25), breaks = seq(.3,1.25, 0.1)) +
  scale_y_continuous(limits = c(9,14), breaks = seq(9, 14, 0.5)) +
  geom_smooth(method = "lm") +
  labs(title = "Sulphates by Alcohol") +
  xlab("Sulphate Level (potassium sulphate - g / dm3)") +
  ylab("Alcohol (% by volume)") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Description Three
This shows little correlation between the two highest positively correlated attributes to wine quality in the dataset, sulphates and alcohol. Since we're ultimately trying to find the attributes which influence the quality of wine and possibly to predict the quality based on these attributes, it's important that the features are not redundant. Redundant attributes lead to a model which overfits predictions.

# Reflections
As wine quality was pretty much a categorical value containing mostly values of 5 or 6, these highly influenced the appearance of the graphs correlating with quality. I was hoping that some of the transforms would give a higher correlation with quality than just the normal attribute, but I didn't see any real evidence of this with the transforms I created.

Some limitations are due to the volume of data. 1599 records is not a large dataset, perhaps I should have chosen the white wines instead. To investigate the data further, I would like to see a larger set. In addition, while the quality measure was a median of three wine experts, I would also like to see the mean in order to show a more continuous variable quality measurement.